The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites

### 2023-06-11 paper/markdown/main_manuscript.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author:
  - Katharina Hüfner:
      institute: [psych2]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true
  - Sophia Vedova:
      institite: [psych2]
  - Piotr Tymoszuk:
      institute: [daas]
  - Philipp Nelles:
      institute: [psych2]
  - Tobias Bruckner: 
      institute: [psych2]    
  - Eberhard Deisenhammer: 
      institute: [psych2, psych1]
  - Jonas Egeter: 
      institute: [psych2]
  - Alexander Egger: 
      institute: [zimcl]
  - Matyas Galffy: 
      institute: [psych2]
  - Johannes Giesinger: 
      institute: [psych2]
  - Jens Lehmann: 
      institute: [psych1]
  - Maria Oberhammer: 
      institute: [psych1]
  - Joachim Rockenschaub: 
      institute: [psych2]
  - Magdalena Sacher: 
      institute: [surg]
  - Bernhard Holzner: 
      institute: [psych2]
  - Johanna M Gostner: 
      institute: [bio]
  - Barbara Sperner-Unterweger: 
      institute: [psych2]

institute:
  - psych2: University Hospital of Psychiatry II, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria. 
  - psych1: University Hospital of Psychiatry I, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
  - surg: Department of Visceral, Transplant and Thoracic Surgery (VTT), Medical University Innsbruck, Innsbruck, Austria. 
  - zimcl: Central Institute of Clinical and Chemical Laboratory Diagnostics, University Hospital of Innsbruck, Innsbruck, Austria.
  - bio:  Insitute of Medical Biochemistry, Medical University Innsbruck, Innsbruck, Austria

@@

Running title: Activity of the kynurenine, serotonin and dopamine metabolic pathways and COVID-19

@@

__Background:__ COVID-19 patients frequently develop symptoms of depression, anxiety and mental stress. 
The serotonin, kynurenine and catecholamine metabolic pathways determine systemic levels of serotonin and dopamine, whose insufficiency was linked to mental health disorders. 
We investigated factors that affect their activity in SARS-CoV-2 infection and recovery.

__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were analyzed. 
Demographic and clinical characteristic, inflammatory markers, SARS-CoV-2 infection, signs of depression and anxiety (HADS), and mental stress (PSS-4) served as explanatory variables. 
Plasma markers of the serotonin and kynurenine (serotonin, kynurenine/tryptophan ratio), and catecholamine pathway activity (dopamine 3-O-sulfate, phenylalanine/tyrosine ratio) were modeled by multi-parameter linear regression.

__Results:__ In the SIMMUN cohort, the inflammatory marker 
neopterin ($\beta$ = `r my_beta(mod_eli$inference$kyn_trp, 'log_neo')`), 
SARS-CoV-2-positivity (`r my_beta(mod_eli$inference$kyn_trp, 'infection', show_ci_txt = FALSE)`), 
mental stress (`r my_beta(mod_eli$inference$kyn_trp, 'pss_stress_score', show_ci_txt = FALSE)`) and age were positively associated with elevated kynurenine/tryptophan ratio. 
Phenylalanine/tyrosine ratio was lower in 
SARS-CoV-2-positive participants (`r my_beta(mod_eli$inference$phe_tyr, 'infection', show_ci_txt = FALSE)`). 
In the INCOV cohort, inflammation markers were associated with decreased serotonin 
(IL6: `r my_beta(incov_neuro$inference$serotonin, 'IL6_INF', show_ci_txt = FALSE)`) and dopamine 3-O-sulfate 
(interferon-gamma: `r my_beta(incov_neuro$inference$dopamine, 'IFNG_INF', show_ci_txt = FALSE)`). 
Serotonin (`r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery', show_ci_txt = FALSE)`) and 
dopamine 3-O-sulfate levels (`r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery', show_ci_txt = FALSE)`) were higher during recovery than in acute SARS-CoV-2 infection.

__Conclusion:__ SARS-CoV-2 infection, inflammation, mental stress and age are the key independent predictors of high kynurenine pathway activity, which decreases systemic serotonin levels by depleting its precursor tryptophan. 
These factors are also associated with suppressed catecholamine pathway activity. 
This lowered serotonin and dopamine availability may contribute to mental health problems in COVID-19 patients.

@@

The immune system and the brain interact at multiple levels [@Dantzer2008]. 
Physical conditions such as infections, cardiovascular disease, frailty, autoimmune illness or malignancy are often accompanied by symptoms of depression, anxiety, mental stress and other psychiatric disorders [@Mitchell2011; @Dickens2001; @Mayerl2020; @Rajan2020; @Renault1987; @Hufner2015; @Hufner2022]. 
Inflammation interferes with key metabolic processes controlling serotonin and dopamine availability, the serotonin, kynurenine and catecholamine pathways, and was proposed as an important link between mental and physical health [@Bonaccorso2002; @Gostner2020; @Valdiglesias2018; @Dantzer2008; @Hufner2015; @Brown2021].

Serotonin is synthesized from the essential aminoacid tryptophan (TRP) via serotonin pathway reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
Most systemic TRP is catabolized by the kynurenine pathway, which limits serotonin pathway activity [@Lukic2022]. 
Its first rate-limiting step, the TRP - kynurenine (KYN) breakdown catalyzed by tryptophan 2,3-dioxygenase (TDO) or indoleamine 2,3-dioxygenases (IDO1, IDO2), controls TRP levels available for serotonin synthesis [@Lukic2022; @Badawy2017]. 
Importantly, IDO1 is stimulated by inflammatory signals [@Badawy2017; @Robinson2003]
and TDO activity is increased by glucocorticoids released upon depression and mental stress [@Knox1951; @Maes1990]. 
Kynurenine pathway metabolites such as KYN or quinolinic acid [QUIN] can additionally pass the blood-brain barrier, interfere e.g. glutamatergic signaling and exert neurotoxicity [@Brown2021; @Schwarcz2012]. 
Hence, the kynurenine pathway, whose activity can be assessed by the ratio of blood KYN to TRP (KYN/TRP), poses a connex of inflammatory and hypothalamic–pituitary–adrenal axis signals, systemic serotonin availability and neural signal transmission [@Badawy2017; @Brown2021; @Schwarcz2012]. 
Consequently, elevated levels of circulating kynurenine pathway metabolites, high KYN/TRP, decreased TRP or lowered serotonin were reported for numerous physical conditions and associated with symptoms of mental disorders [@Cervenka2017; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Capuron2011].

Dopamine synthesis via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH). 
PAH, TH and the serotonin pathway enzyme TPH rely on tetrahydrobiopterin (BH4) as an essential cofactor [@Meiser2013; @Neurauter2008]. 
During inflammation, BH4 is oxidized to its inactive derivative and consumed in a competitor reaction of nitric oxide formation [@Rahimian2022a; @Neurauter2008]. 
As a result of such functional BH4 deficiency, reflected e.g. by an elevated blood PHE to TYR ratio, biosynthesis of dopamine and serotonin is suppressed [@Geisler2013; @Vancassel2018; @Capuron2011]. 
Accordingly, changes in PHE and TYR levels have been reported in different physical disorders such as cancer, infections and inflammatory conditions and associated with depression and anxiety [@Vancassel2018; @Geisler2013; @Hufner2015; @Capuron2011].

SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
Apart from sustained physical disability, psychiatric disorders in millions of patients amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022]. 
Increased kynurenine pathway activity suggestive of lowered systemic serotonin availability has recently been identified in acute COVID-19 and during COVID-19 recovery and associated with infection severity [@Santiago-Mujika2022; @Lionetto2021; @Bizjak2022; @Dewulf2022]. 
Theoretical considerations [@Bower2022] and preliminary experimental data suggest that highly active kynurenine pathway along with suppression of serotonin and catecholamine biosynthesis may contribute to symptoms of mental stress, anxiety and depression in COVID-19 patients [@Matits2023; @Kucukkarapinar2022].

Still, clinical and experimental evidence for interaction of inflammation, mental health and systemic availability of serotonin and dopamine in COVID-19 is scarce. 
To address this question, we explored effects of demographic, clinical, psychometric (anxiety, depression, mental stress), inflammation- and SARS-CoV-2-related factors on markers of systemic activity of the serotonin, kynurenine and catecholamine pathways in two cross-sectional cohorts of SARS-CoV-2 patients: the SIMMUN cohort and the published longitudinal INCOV study [@Su2022].

@@

The SIMMUN data were collected during a single on-site visit at median 139 days after the SARS-CoV-2 PCR (interquartile range: 119 - 157) or extracted from electronic patient records (__Supplementary Table S1__). 
Demographic and clinical characteristic included age, sex, body mass index (BMI), mental disorders diagnosed by a physician, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR. 
Inflammatory markers, plasma neopterin concentrations (NEO) measured by ELISA [@Werner1990] and neutrophil/lymphocyte ratio (NLR) were determined by a certified laboratory at the University Hospital of Innsbruck. 
Plasma TRP, KYN, PHE and TYR were determined by high-performance liquid chromatography [@Widner1997; @Neurauter2008]. 
Immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) was quantified by ELISA [@Deisenhammer2021]. 
Laboratory measurements beyond the detection limits were substituted with the respective detection limit value. 
Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Clinically relevant symptoms of anxiety and depression were defined as $\geq$ 8 points in the hospital anxiety and depression scale (HADS) [@Bjelland2002; @Zigmond1983].

@@

Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and liquid chromatography/tandem mass spectrometry (Metabolon, USA), respectively, and expressed as age- and sex-adjusted, log~2~-transformed concentrations [@Su2022; @Su2020]. 
Markers of the serotonin (serotonin, TRP) [@Lukic2022], kynurenine (KYN, kynurenine/tryptophan ratio [KYN/TRP], QUIN) [@Badawy2017] and catecholamine pathway activity (PHE, TYR, phenylalanine/tyrosine ratio [PHE/TYR] and dopamine 3-O-sulfate [DA sulfate]) [@Meiser2013; @Goldstein1999; @Suominen2013], cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]), age, sex and BMI were analyzed (__Supplementary Table S3__).

Plasma samples were collected in during in acute and sub-acute infection, and recovery at median 
`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`, 
`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`), 
`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR, respectively (__Supplementary Table S4__).

@@

The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma markers of kynurenine and catecholamine pathway activity (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) by multi-parameter modeling in the SIMMUN cohort. 
The second endpoint was to investigate how markers of systemic serotonin and dopamine availability (DA sulfate), are influenced by age, sex, body mass, inflammatory cytokines, timepoint of SARS-CoV-2 infection, neurotransmitter precursor amino acids (TRP, PHE, TYR) and kynurenine pathway metabolites (KYN, QUIN). 
This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Supplementary Figure S1__).

@@

Responses and explanatory variables were normalized prior to modeling. 
In the SIMMUN cohort, effects of age, sex, body mass class, physical and psychiatric disorders, BMI, smoking and alcohol consumption, inflammation markers (NEO, NLR), SARS-CoV-2 infection, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress (PSS-4) on pathway activity markers were assessed by multi-parameter linear regression with backward elimination. 
In the INCOV cohort, effects of age, sex, BMI, cytokine markers of inflammation (IL6, IL10, TNF, IFNG), SARS-CoV-2 infection timepoint (uninfected, acute, sub-acute, recovery), serotonin and catecholamine pathway metabolites (TRP, PHE, TYR) and kynurenine pathway metabolites (QUIN, KYN) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression [@Huber2011]. 
Explanatory performance, reproducibility and proper parameterization of the multi-parameter models was investigated by R^2^ and root mean square error (RMSE) in 10-fold cross-validation. 
Differences in cytokines and metabolites between SARS-CoV-2 infection timepoints in the INCOV collective were investigated by robust linear modeling with uninfected subset or acute infection serving as baselines.

@@

The excluded individuals were characterized by more frequent mental disorders and clinically relevant symptoms of depression and anxiety (HADS), higher scores of mental stress (PSS-4) and less frequent SARS-CoV-2 infections than the analyzed participants (__Supplementary Table S2__). 

@@

The psychometric tools for assessment of perceived mental stress (PSS-4), depression and anxiety (HADS) displayed good-to-excellent internal consistency 

@@

(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_dpr_score') %>% .[2]`%), yet this effect was no statistically significant. 

@@

As expected, titer of antibodies against the S1/S2 SARS-CoV-2 protein (anti-RBD IgG) was significantly higher in the SARS-CoV-2-positive subset 
(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$anti_rbd %>% median %>% signif(2)` AU) as compared with uninfected individuals 
(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% .$anti_rbd %>% median %>% signif(2)` AU, 
`r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$significance`, 
effect size: `r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$eff_size`). 

@@

In the SIMMUN dataset, we initially searched for predictors of activity of the kynurenine and catecholamine pathways. 
TRP, KYN and KYN/TRP served as markers of the kynurenine pathway [@Badawy2017], and PHE, TYR and PHE/TYR were investigated as markers of the catecholamine pathway activity [@Meiser2013]. 
The candidate explanatory variables were age, sex, BMI, the presence of physical and mental disorders, smoking, alcohol consumption, SARS-CoV-2 infection, anti-SARS-CoV-2 antibody titer, NEO and NLR as inflammatory markers, perceived mental stress, and symptoms of depression and anxiety (__Supplementary Table S1__).

Meaningful multi-parameter models optimized by backward elimination could be established for TRP, KYN, KYN/TRP, TYR and PHE/TYR. The PHE model had a negligible explanatory performance, proved non-significant as compared with the intercept-only null model and was hence not further analyzed  (__Supplementary Table S6__). 
The remaining models were characterized by good reproducibility and proper parameterization as indicated by comparable fit errors (RMSE) in the training dataset and cross-validation. 
The KYN and KYN/TRP ratio models had the best, moderate-to-substantial explanatory performance measured by cross-validated R^2^ of 
`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn') %>% .$estimate %>% signif(2)` and 
`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn_trp') %>% .$estimate %>% signif(2)`, respectively. 
The cross-validated R^2^ metric values for the TRP, TYR and PHE/TYR models ranged from 
`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1]`% to 
`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[2]` indicative of weak-to-moderate explanatory value (__Supplementary Figure S3__).

>>>>

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author:
  - Katharina Hüfner:
      institute: [psych2]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true
  - Sophia Vedova:
      institite: [psych2]
  - Piotr Tymoszuk:
      institute: [daas]
  - Philipp Nelles:
      institute: [psych2]
  - Tobias Bruckner: 
      institute: [psych2]    
  - Eberhard A. Deisenhammer: 
      institute: [psych2, psych1]
  - Jonas Egeter: 
      institute: [psych2]
  - Matyas Galffy: 
      institute: [psych2]
  - Johannes Giesinger: 
      institute: [psych2]
  - Jens Lehmann: 
      institute: [psych2]
  - Maria Oberhammer: 
      institute: [psych1]
  - Joachim Rockenschaub: 
      institute: [psych2]
  - Magdalena Sacher: 
      institute: [surg]
  - Bernhard Holzner: 
      institute: [psych2]
  - Johanna M Gostner: 
      institute: [bio]
  - Barbara Sperner-Unterweger: 
      institute: [psych2]

institute:
  - psych2: University Hospital of Psychiatry II, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria. 
  - psych1: University Hospital of Psychiatry I, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
  - surg: Department of Visceral, Transplant and Thoracic Surgery (VTT), Medical University Innsbruck, Innsbruck, Austria. 
  - bio:  Insitute of Medical Biochemistry, Medical University Innsbruck, Innsbruck, Austria

@@

Running title: Activity of the serotonin, kynurenine, and catecholamine metabolic pathways and COVID-19

@@

__Background:__ Increased prevalence of mental disorders following COVID-19 has been described. 
The kynurenine and catecholamine metabolic pathways are modulated by inflammation and can affect systemic levels of serotonin and dopamine. 
Their activity may hence link physical disorders with mental health. 
We investigated factors that affect kynurenine and catecholamine pathway activity in SARS-CoV-2 infection and recovery.

__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were analyzed. 
Demographic and clinical characteristic, inflammatory markers, SARS-CoV-2 infection, symptoms of depression and anxiety (HADS), and mental stress (PSS-4) served as explanatory variables. 
Blood serotonin and markers of the kynurenine (kynurenine/tryptophan ratio), and catecholamine pathway activity (dopamine 3-O-sulfate, phenylalanine/tyrosine ratio) were modeled by multi-parameter linear regression.

__Results:__ In the SIMMUN cohort, the inflammatory marker 
neopterin ($\beta$ = `r my_beta(mod_eli$inference$kyn_trp, 'log_neo')`), 
SARS-CoV-2-positivity (`r my_beta(mod_eli$inference$kyn_trp, 'infection', show_ci_txt = FALSE)`), 
mental stress (`r my_beta(mod_eli$inference$kyn_trp, 'pss_stress_score', show_ci_txt = FALSE)`) and age were positively associated with elevated kynurenine/tryptophan ratio. 
Phenylalanine/tyrosine ratio was lower in 
SARS-CoV-2-positive than SARS-CoV-2-negative participants (`r my_beta(mod_eli$inference$phe_tyr, 'infection', show_ci_txt = FALSE)`). 
In the INCOV cohort, inflammation markers were associated with decreased serotonin 
(IL6: `r my_beta(incov_neuro$inference$serotonin, 'IL6_INF', show_ci_txt = FALSE)`) and dopamine 3-O-sulfate 
(interferon-gamma: `r my_beta(incov_neuro$inference$dopamine, 'IFNG_INF', show_ci_txt = FALSE)`). 
Serotonin (`r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery', show_ci_txt = FALSE)`) and 
dopamine 3-O-sulfate levels (`r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery', show_ci_txt = FALSE)`) were higher during recovery than in acute SARS-CoV-2 infection.

__Conclusion:__ 
SARS-CoV-2 infection, inflammation, age and mental stress are the key independent predictors of high kynurenine pathway activity, which may reduce systemic serotonin availability. 
The catecholamine pathway was suppressed in SARS-CoV-2 infection. 
Altered activity of these pathways may contribute to mental disorder symptoms following COVID-19.

@@

The immune system and the brain interact at multiple levels [@Dantzer2008]. 
Physical disorders such as infections, cardiovascular disease, frailty, autoimmune illness or malignancy are often accompanied by symptoms of depression, anxiety and mental stress [@Mitchell2011; @Dickens2001; @Mayerl2020; @Rajan2020; @Renault1987; @Hufner2015; @Hufner2022]. 
Inflammation interferes with key metabolic processes such as the kynurenine and catecholamine pathway, which can control serotonin and dopamine availability, respectively. 
Inflammation was thus proposed as an important link between mental and physical health [@Bonaccorso2002; @Gostner2020; @Valdiglesias2018; @Dantzer2008; @Hufner2015; @Brown2021].

Serotonin is synthesized from the essential aminoacid tryptophan (TRP) in reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
Most systemic TRP is catabolized along the kynurenine pathway. 
The conversion of TRP to kynurenine (KYN) by tryptophan 2,3-dioxygenase (TDO) or indoleamine 2,3-dioxygenases (IDO1, IDO2) is the first, rate-limiting step of the kynurenine pathway and may thereby control TRP levels subsequently available for serotonin synthesis [@Lukic2022; @Badawy2017]. 
Importantly, IDO1 is stimulated by inflammatory signals [@Badawy2017; @Robinson2003]
and TDO activity is increased by glucocorticoids which are also released as a reaction to mental stress [@Knox1951; @Maes1990]. 
Hence, activity of IDO and TDO, which can be assessed by the ratio of blood KYN to TRP (KYN/TRP), is modulated by inflammatory and hypothalamic–pituitary–adrenal axis signals.
Additionally, kynurenine pathway metabolites such as KYN or quinolinic acid [QUIN] can additionally pass the blood-brain barrier, interfere e.g. glutamatergic signaling and exert neurotoxicity [@Badawy2017; @Brown2021; @Schwarcz2012]. 
Consequently, elevated concentrations of circulating kynurenine pathway metabolites, high KYN/TRP, decreased TRP as well as lowered levels of associated neurotransmitter serotonin were reported for numerous physical disorders and associated with symptoms of mental disorders [@Cervenka2017; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Capuron2011].

The synthesis of catecholamine neurotransmitters such has dopamine via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH). 
PAH, TH and the serotonin-synthesizing TPH enzyme rely on tetrahydrobiopterin (BH4) as an essential cofactor [@Meiser2013; @Neurauter2008]. 
During inflammation, BH4 is oxidized to its inactive derivative and consumed in a competitor reaction of nitric oxide formation [@Rahimian2022a; @Neurauter2008]. 
As a result of such functional BH4 deficiency, reflected e.g. by an elevated blood PHE to TYR ratio, biosynthesis of dopamine and serotonin is reduced [@Geisler2013; @Vancassel2018; @Capuron2011]. 
Accordingly, changes in PHE and TYR levels have been reported in different physical disorders such as cancer, infections and inflammatory conditions and associated with depression and anxiety [@Vancassel2018; @Geisler2013; @Hufner2015; @Capuron2011].

SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
Apart from sustained physical disability, mental disorders in millions of patients amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022]. 
Increased kynurenine pathway activity has recently been identified in acute COVID-19 and during COVID-19 recovery and associated with infection severity [@Santiago-Mujika2022; @Lionetto2021; @Bizjak2022; @Dewulf2022]. 
Theoretical considerations [@Bower2022] and preliminary experimental data suggest that a highly active kynurenine pathway as well as reduced biosynthesis of serotonin and catecholamine neurotransmitters may contribute to symptoms of mental stress, anxiety and depression in COVID-19 patients [@Matits2023; @Kucukkarapinar2022].

Clinical and experimental evidence for an interaction of mental health with inflammation, kynurenine and catecholamine pathway activity and systemic availability of serotonin and dopamine in COVID-19 is scarce. 
To address this question, we analyzed the effects of demographic, clinical, psychometric (anxiety, depression, mental stress), inflammation- and SARS-CoV-2-related factors on the kynurenine pathway activity, serotonin levels as well as the catecholamine pathway in two cohorts of SARS-CoV-2 patients: the cross-sectional SIMMUN cohort and the published longitudinal INCOV cohort [@Su2022].

@@

The SIMMUN data were collected during a single on-site visit at median 139 days after the SARS-CoV-2 PCR (interquartile range: 119 - 157) and extracted from electronic patient records (__Supplementary Table S1__). 
Demographic and clinical characteristics included age, sex, body mass index (BMI), mental disorders diagnosed by a physician, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR. 
Neutrophil/lymphocyte ratio (NLR) and serum concentration of neopterin (NEO) served as inflammatory markers. 
NLR was determined by a certified laboratory at the University Hospital of Innsbruck. 
NEO was measured measured by enzyme-linked immunosorbent assay (BRAHMS Diagnostics, Berlin, Germany). 
Serum TRP, KYN, PHE and TYR were determined by high-performance liquid chromatography [@Widner1997; @Neurauter2008]. 
Immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) was quantified by ELISA [@Deisenhammer2021]. 
Laboratory measurements beyond the detection limits were substituted with the respective detection limit value. 
Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Clinically relevant symptoms of anxiety and depression were defined as $\geq$ 8 points in the hospital anxiety and depression scale (HADS) [@Bjelland2002; @Zigmond1983].

@@

Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and liquid chromatography/tandem mass spectrometry (Metabolon, USA), respectively, and expressed as age- and sex-adjusted, log~2~-transformed concentrations [@Su2022; @Su2020]. 
Serotonin, metabolites of the kynurenine (TRP, KYN, QUIN) [@Lukic2022; @Badawy2017] and catecholamine pathway (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]) [@Meiser2013; @Goldstein1999; @Suominen2013], cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]), age, sex and BMI were analyzed (__Supplementary Table S3__).

Plasma samples were collected during acute and sub-acute infection, and recovery at median 
`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`, 
`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`, 
`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR, respectively (__Supplementary Table S4__).

@@

The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing serum markers of kynurenine and catecholamine pathway activity (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) by multi-parameter modeling in the SIMMUN cohort. 
The second endpoint was to investigate how systemic serotonin and DA dopamine availability (DA sulfate), are influenced by age, sex, BMI, cytokine markers of inflammation, timepoint of SARS-CoV-2 infection, kynurenine pathway (TRP, KYN, QUIN) and catecholamine metabolite levels (PHE, TYR). 
This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Supplementary Figure S1__).

@@

Responses and numeric explanatory variables were normalized prior to modeling. 
In the SIMMUN cohort, effects of age, sex, body mass class, physical and psychiatric disorders, BMI, smoking and alcohol consumption, inflammation markers (NEO, NLR), SARS-CoV-2 infection, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress (PSS-4) on kynurenine and catecholamine pathway activity markers were assessed by multi-parameter linear regression with backward elimination. 
In the INCOV cohort, effects of age, sex, BMI, cytokine markers of inflammation (IL6, IL10, TNF, IFNG), SARS-CoV-2 infection status and timepoint (uninfected, acute, sub-acute, recovery), kynurenine (TRP, KYN, QUIN) and catecholamine pathway metabolites (PHE, TYR) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression [@Huber2011]. 
Explanatory performance, reproducibility and proper parameterization of the multi-parameter models was investigated by R^2^ and root mean square error (RMSE) in 10-fold cross-validation. 
Differences in cytokines and metabolites between SARS-CoV-2 infection timepoints in the INCOV collective were investigated by robust linear modeling with uninfected subset or acute infection serving as baselines.

@@

The excluded individuals were characterized by more frequent mental disorders and more frequent clinically relevant symptoms of depression and anxiety (HADS), higher scores of mental stress (PSS-4) and less frequent SARS-CoV-2 infections than the analyzed participants (__Supplementary Table S2__). 

@@

The psychometric tools for assessment of perceived mental stress (PSS-4), and depression and anxiety (HADS) displayed good-to-excellent internal consistency 

@@

(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_dpr_score') %>% .[2]`%), yet this effect was not statistically significant. 

@@

@@

In the SIMMUN dataset, we initially searched for predictors of activity of the kynurenine and catecholamine pathways. 
TRP, KYN and KYN/TRP served as markers of the kynurenine pathway, and PHE, TYR and PHE/TYR were investigated as markers of the catecholamine pathway activity. 
The candidate explanatory variables were age, sex, BMI, the presence of physical and mental disorders, smoking, alcohol consumption, SARS-CoV-2 infection, anti-SARS-CoV-2 antibody titer, NEO and NLR as inflammatory markers, perceived mental stress, and symptoms of depression and anxiety (__Supplementary Table S1__).

Meaningful multi-parameter models optimized by backward elimination could be established for TRP, KYN, KYN/TRP, TYR and PHE/TYR. The PHE model had a negligible explanatory performance, proved non-significant as compared with the intercept-only null model and was hence not further analyzed  (__Supplementary Table S6__). 
The remaining models were characterized by good reproducibility and proper parameterization as indicated by comparable fit errors (RMSE) in the training dataset and cross-validation. 
The KYN and KYN/TRP ratio models had the best, moderate-to-substantial explanatory performance measured by cross-validated R^2^ of 
`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn') %>% .$estimate %>% signif(2)` and 
`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn_trp') %>% .$estimate %>% signif(2)`, respectively. 
The cross-validated R^2^ metric values for the TRP, TYR and PHE/TYR models ranged from 
`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1]` to 
`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[2]` indicative of weak-to-moderate explanatory value (__Supplementary Figure S3__).
